A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-05-21 paper/writeup.tex
.

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tscompare}{\texttt{tscompare}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\tsdate}{\texttt{tsdate}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\pluseq}{\mathrel{{+}{=}}}

@@

A common measure of disagreement between ARGs, 
first proposed by~\cite{kuhner2015assessing}, 
is to use a weighted average Robinson-Foulds (RF) distance.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(n) = \left\{ k : \exists\ n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(n)} (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average RF distance (averaged over locations in the genome) \revpoint{1}{2}
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average RF distance allows a different node to match
on each tree.
The other differences are that \emph{average} RF distance
normalizes by sequence length, and is symmetrized.
The RF distance between two trees
was defined by \citet{robinson1981comparison}
% JK - change edge->branch here to avoid confusion with tskit/ARG edge
to be the minimum number of branch contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges 
that induce different splits on the labels).
A similar metric on ARGs could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

The implementation of this method in \tscompare{}
additionally produces relative values:
the ARG RF value (non-matched span relative to $\T_1$)
and true proportion represented (inverse matched span relative to $\T_2$).
The ARG RF (which we call ``ARF'') is defined to be the matched span proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\operatorname{ARF}(\T_1,\T_2)
 	=1-\frac{\match(\T_1,\T_2)}{\Vert \T_1\Vert} ,
\end{align}
and so if $\T_2$ represents the truth, is analogous to a false positive rate.
The \emph{true proportion represented} (TPR) is the inverse matched span between 
two trees relative to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\operatorname{TPR}(\T_1,\T_2) = \frac{\invmatch(\T_1,\T_2)}
	{\Vert \T_2\Vert} ,
\end{align}
and is analogous to statistical power.
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original matched span and inverse matched span, whose units are length of spans.
We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.\revpoint{1}{4}

@@

\paragraph{Metrics on ARGs} \revpoint{1}{5}
Neither the matched span or non-matched span of $\T_1$ in $\T_2$ are metrics in the mathematical sense
(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
This is by design: in practice it is not possible to infer all aspects
of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
and so we wanted to quantify
``How much of the true relationships does this ARG represent?''
However, it is worth noting that the symmetrized version of non-matched span
$$\overline{\nmatch}(\T_1,\T_2) = \nmatch(\T_1,\T_2) + \nmatch(\T_2, \T_1),$$
 is a metric.
 To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$, called $\varphi\colon N\to N$, where $N$ is the shared set of nodes.
 View both ARGs as a subset of the space $N\times N\times [0,L)$ so that elements in the ARG are tuples $(n, \varphi(n), A)$ where $A=\cup_k [a_{k-1}, a_k)$ represents the portion of the genome where $n$ exists in the ARG.
 Then for each node $n$ in $N$ its representatives in $\T_1$ and $\T_2$ can differ with respect to the portion of the genome where $n$ exists, ie. $A_{\T_1}$ and $A_{\T_2}$ may not be equal.
 Then the relative difference (as sets) of the two ARGs, $\T_1\setminus \T_2$, is the collection of tuples $(n, \varphi(n), A_{\T_1}\setminus A_{\T_2})$. This is the portion of \emph{non-matched spans} between $n$ and $\varphi(n)$.
%To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
%and view each ARG as a subset of the space $[0,L) \times N \times N$,
%where $N$ is the shared set of nodes.
Thus, the measure of non-matched spans
%dissimilarity 
is the Lebesgue measure of the relative difference of the two sets:
$|\T_1 \setminus \T_2|$,
and so the symmetrized version is the measure of the symmetric difference
$|\T_1\Delta\T_2|$,
which is well-known to be a metric \citep{rudin1976principles}. 
If the two ARGs have the same number of nodes,
we can consider all bijections between their nodes.
The symmetric difference between $\T_1$ and $\T_2$,
related through each bijection is a metric.
Then the minimum over all such metrics will still be a metric
since the minimum over a finite number of metrics is also a metric.
This also extends to two tree sequences with different numbers of nodes,
$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $\T_1\to\T_2$ and $\T_2\to\T_1$.

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tscompare}{\texttt{tscompare}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\tsdate}{\texttt{tsdate}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\pluseq}{\mathrel{{+}{=}}}
\newcommand{\arf}\operatorname{ARF}
\newcommand{\tpr}\operatorname{TPR}

@@

A common measure of disagreement between ARGs, 
first proposed by~\cite{kuhner2015assessing}, 
is to use a weighted average Robinson-Foulds (RF) distance.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(n) = \left\{ k : \exists\ n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(n)} (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average RF distance (averaged over locations in the genome) \revpoint{1}{2}
is
\begin{align*}
    \frac{1}{L} \left( \|\T_1\| + \|\T_2\|
        - \sum_{n_1 \in N_1} m'(n_1, \T_2) - \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
(This holds because the average number of nodes per local tree in $\T_1$ is $\|\T_1\|/L$,
so the difference gives the average number of non-matching nodes.)
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average RF distance allows a different node to match
on each tree.
The other differences are that \emph{average} RF distance
normalizes by sequence length, and is symmetrized.
The RF distance between two trees
was defined by \citet{robinson1981comparison}
to be the minimum number of branch contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of branches 
that induce different splits on the labels).
A similar metric on ARGs could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

The implementation of this method in \tscompare{}
additionally produces relative values:
the ARG RF value (non-matched span relative to $\T_1$)
and true proportion represented (inverse matched span relative to $\T_2$).
The ARG RF (which we call ``ARF'') is defined to be the matched span proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\operatorname{ARF}(\T_1,\T_2)
 	=1-\frac{\match(\T_1,\T_2)}{\Vert \T_1\Vert} ,
\end{align}
and so if $\T_2$ represents the truth, is analogous to a false positive rate.
The \emph{true proportion represented} (TPR) is the inverse matched span between 
two trees relative to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\operatorname{TPR}(\T_1,\T_2) = \frac{\invmatch(\T_1,\T_2)}
	{\Vert \T_2\Vert} ,
\end{align}
and is analogous to statistical power.
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original matched span and inverse matched span, whose units are length of spans.
We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.

@@

\paragraph{Metrics on ARGs} \revpoint{1}{5}
Neither the matched span or non-matched span of $\T_1$ in $\T_2$ are metrics in the mathematical sense
(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
This is by design: in practice it is not possible to infer all aspects
of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
and so we wanted to quantify
``How much of the true relationships does this ARG represent?''
Indeed, even a symmetrized version of non-matched span is not a metric:
for instance, if $\T_2$ is produced from $\T_1$ by adding a new, entirely unary node,
then $\nmatch(\T_1,\T_2) = \nmatch(\T_2,\T_1) = 0$.
This is again by design: in general, edges in ARGs represent transmission of genetic material
through many ancestors, and so addition of entirely unary ancestors on an edge
is (arguably) not wrong.

However, it is worth noting that the symmetrized version of non-matched span
$$\overline{\nmatch}(\T_1,\T_2) = \nmatch(\T_1,\T_2) + \nmatch(\T_2, \T_1)$$
is a metric.
To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
and so can treat them as having the same set of samples $S$ and nodes $N$
(with $S \subset N$).
To each ARG $\T_i$ corresponds a mapping $\mathcal{D}_i : [0, L) \times N \to \{0,1\}^S$,
such that $\mathcal{D}_i(u,n)$ gives the set of samples that descend from $n$ at position $u$
along the genome,
and hence a set $P_i = \{(u,n,\mathcal{D}_i(u,n)): u \in [0,L), n \in N\}$.
The non-matched span $\nmatch(\T_1,\T_2)$
is the Lebesgue measure of the relative difference of the two sets:
$|P_1 \setminus P_2|$,
and so the symmetrized version is the measure of the symmetric difference
$|\T_1\Delta\T_2|$,
which is well-known to be a metric \citep{rudin1976principles}. 
If the two ARGs have the same number of nodes,
we can consider all bijections between their nodes.
The symmetric difference between $\T_1$ and $\T_2$,
related through each bijection is a metric.
Then the minimum over all such metrics will still be a metric
since the minimum over a finite number of metrics is also a metric.
This also extends to two tree sequences with different numbers of nodes,
since we can take the minimum over all possible matchings $\T_1\to\T_2$ and $\T_2\to\T_1$.
